Generalized Plackett-Luce Likelihoods

### 2022-08-09 inst/jss_hyper3.Rnw
change insect example to chorioamnion dataset from wilcox.Rd

The ideas presented above can easily be extended to arbitrarily large
numbers of competitors.  Given an order statistic of the type
considered above, the Mann-Whitney-Wilcoxon
test~\citep{mann1947,wilcoxon1945} assesses a null of identity of
underlying distributions.  Consider the fly spray dataset presented by
\citeauthor{wilcoxon1945}:

<<defineAB>>=
A <- c(68,68,59,72,64,67,70,74)
B <- c(60,67,61,62,67,63,56,58)
@

Above we see the percentage mortality of two preparations, A and B,
each applied eight times.  We may easily convert this dataset to a
rank order statistic:

<<makeABorderstatistic>>=
names(A) <- rep("a",length(A))
names(B) <- rep("b",length(B))
sort(c(A,B))
names(sort(c(A,B)))
@

And thence to a likelihood function:

<<definehyper3AB,cache=TRUE>>=
(Hab <- ordervec2supp3(names(sort(c(A,B)))))
@

Package idiom for testing the null of equality is straightforward:

<<testABforequality,cache=TRUE>>=
equalp.test(Hab)
@

Compare \citeauthor{wilcoxon1945}'s $p$-value of $1.77\%$.  However,
observe that the \code{hyper3} likelihood approach gives more
information than Wilcoxon's analysis: Firstly, we see from the
evaluate that the probability of a single application of preparation A
killing more flies than B is about 15\%.  Further, we may plot a
support curve for this dataset as above:

\begin{figure}[htbp]
  \begin{center}
<<plotwilcoxlike,cache=TRUE>>=
a <- seq(from=0.02,to=0.5,len=50)
L <- sapply(a,function(p){loglik(p,Hab)})
plot(a,L-max(L),type='b')
abline(h=c(0,-2))
abline(v=0.1558)
@
\caption{A\label{wilcox_likelihood} support for the Bradley-Terry
strength $p_a$ of preparation A~\citep{wilcoxon1945}.  The evaluate of
0.1558 is shown; and the two-units-of support credible interval, which
excludes $p_a=0.5$}
\end{center}
\end{figure}

@@

The ideas presented above are easily extended to multiple types of
competitors.  Consider the following table, drawn from the men's
javelin, 2020 Olympics:

@@

This Chopra threw 87.03m on his first throw, 87.58m on his second, and
so on.  No-throws, ignored here, are indicated with a \code{X}.  We
may convert this to a named vector with elements being the throw
distances, and names being the competitors, using
`attemptstable2supp3()`:

@@

In Formula 1 motor racing, the constructors' championship takes place
between {\em manufacturers} of racing cars (compare the drivers'
championship).  Here, we consider the constructor to be the object of
inference.  Each constructor typically fields two cars, each of which
separately accumulates ranking-based points at each venue.  Here we
use generalized Plackett-Luce to assess the contructors' performance.
The following table shows rankings for the first 9 venues of the 2021
season:

@@

Above, we see that Mercedes (``\code{Merc}'') came first and third at
Bahrain (\code{BHR}, and at Emilia Romagna (\code{EMI} came second and
retired (full details of the notation and conventions is given in the
package at \code{constructor.Rd}).  The identity of the driver is not
admissible information and indeed may change during a season.
Alternatively, we may regard the driver and the constructor as a joint
entity, with the constructor's ability to attract and retain a skilled
driver being part of the object of inference.

The associated generalized Plackett-Luce \code{hyper3} object is as follows:

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{constructor_2021}
\end{alltt}
\begin{verbatim}
## log( (AMM=1)^38 * (AMM=1, AR=1)^-1 * (AMM=1, AR=1, ARRF=1, ATH=1, HF=2,
## MM=1, Merc=1, WM=2)^-1 * (AMM=1, AR=1, ARRF=1, ATH=1, HF=2, MM=1,
## WM=2)^-2 * (AMM=1, AR=1, ARRF=1, ATH=1, HF=2, MM=2, Merc=1, WM=2)^-1 *
## (AMM=1, AR=1, ARRF=1, ATH=1, HF=2, RBRH=1, WM=1)^-1 * (AMM=1, AR=1,
## ARRF=1, ATH=1, HF=2, RBRH=1, WM=2)^-1 * (AMM=1, AR=1, ARRF=1, ATH=1,
## HF=2, WM=2)^-1 * (AMM=1, AR=1, ARRF=1, ATH=2, Ferrari=1, HF=2, WM=1)^-1
## * (AMM=1, AR=1, ARRF=1, HF=2, Merc=2, RBRH=1, WM=2)^-1 * (AMM=1, AR=1,
## ARRF=1, HF=2, WM=1)^-1 * (AMM=1, AR=1, ARRF=2, ATH=1, Ferrari=1, HF=2,
## MM=1, Merc=1, RBRH=1)^-1 * (AMM=1, AR=1, ARRF=2, ATH=1, Ferrari=1,
## HF=2, MM=1, Merc=1, RBRH=1, WM=1)^-1 * (AMM=1, AR=1, ARRF=2, ATH=1,
## Ferrari=1, HF=2, MM=1, Merc=2, RBRH=1, WM=2)^-1 * (AMM=1, AR=1, ARRF=2,
...
\end{verbatim}
\end{kframe}
\end{knitrout}

>>>>

The ideas presented above can easily be extended to arbitrarily large
numbers of competitors.  Given an order statistic of the type
considered above, the Mann-Whitney-Wilcoxon
test~\citep{mann1947,wilcoxon1945} assesses a null of identity of
underlying distributions.  Consider the chorioamnion dataset at
\code{wilcox.test.Rd}:

<<define_xy_wilcox>>=
x <- c(0.80, 0.83, 1.89, 1.04, 1.45, 1.38, 1.91, 1.64, 0.73, 1.46)
y <- c(1.15, 0.88, 0.90, 0.74, 1.21)
@

Here we see permeability constants of the human chorioamnion at term
(\code{x}) and between 12 to 26 weeks gestational age (\code{y}).  The
order statistic is straightforward to calculate:

<<hyper3osdef,cache=TRUE>>=
names(x) <- rep("x",length(x))
names(y) <- rep("y",length(y))
(os <- names(sort(c(x,y))))
@

Then object \code{os} is converted to a \code{hyper3} object, again
with \code{ordervec2supp3()}, which may be assessed using the Method
of Support:

<<hyper3xytest,cache=TRUE>>=
Hxy <- ordervec2supp3(os)
equalp.test(Hxy)
@

Above, we use generic function \code{equalp.test()} to test the null
that the permeability of the two groups both have Bradley-Terry
strength of $0.5$.  We see a p-value of about 0.09; compare 0.25 from
\code{wilcox.test()}.  However, observe that the \code{hyper3}
likelihood approach gives more information than Wilcoxon's analysis:
Firstly, we see that the maximum likelihood estimate for the
Bradley-Terry strength of \code{x} is about 0.24, considerably less
than the null of 0.5; further, we may plot a support curve for this
dataset, given in Figure~\ref{wilcox_likelihood}.

\begin{figure}[htbp]
  \begin{center}
<<plotwilcoxlike,cache=TRUE>>=
a <- seq(from=0.02,to=0.8,len=40)
L <- sapply(a,function(p){loglik(p,Hxy)})
plot(a,L-max(L),type='b')
abline(h=c(0,-2))
abline(v=c(0.24))
abline(v=c(0.5),lty=2)
@
\caption{A\label{wilcox_likelihood} support for the Bradley-Terry
strength $p_a$ of permeability at term.  The evaluate of 0.24 is
shown; and the two-units-of support credible interval, which does not
exclude $H_0\colon p_a=0.5$ (dotted line), is also shown}
\end{center}
\end{figure}

@@

The ideas presented above may be extended to more than two types of
competitors.  Consider the following table, drawn from the men's
javelin, 2020 Olympics:

@@

This Chopra threw 87.03m on his first throw, 87.58m on his second, and
so on.  No-throws, ignored here, are indicated with a \code{X}.  We
may convert this to a named vector with elements being the throw
distances, and names being the competitors, using
\code{attemptstable2supp3()}:

@@

In Formula 1 motor racing, the constructors' championship takes place
between {\em manufacturers} of racing cars (compare the drivers'
championship).  Here, we consider the constructor to be the object of
inference.  Each constructor typically fields two cars, each of which
separately accumulates ranking-based points at each venue.  Here we
use generalized Plackett-Luce to assess the contructors' performance.
The following table, included in the \code{hyper2} package as a
dataset, shows rankings for the first 9 venues of the 2021 season:

@@

Above, we see that Mercedes (``\code{Merc}'') came first and third at
Bahrain (\code{BHR}, and at Emilia Romagna (\code{EMI} came second and
retired (full details of the notation and conventions is given in the
package at \code{constructor.Rd}).  The identity of the driver is
viewed as inadmissible information and indeed may change during a
season.  Alternatively, we may regard the driver and the constructor
as a joint entity, with the constructor's ability to attract and
retain a skilled driver being part of the object of inference.  The
associated generalized Plackett-Luce \code{hyper3} object is easily
constructed using package idiom, in this case
\code{ordertable2supp3()}, and we may use this to assess the
Plackett-Luce strengths of the constructors:

<<maxpconstructor2021, cache=TRUE>>=
const2020 <- ordertable2supp3(constructor_2020_table)
const2021 <- ordertable2supp3(constructor_2021_table)
maxp(const2020,n=1)
maxp(const2021,n=1)
@

Above, we see the strength of Mercedes falling from about 0.38 in 2020
to less than 0.20 in 2021 and it is natural to wonder whether this can
be ascribed to random variation.  Observe that testing such a
hypothesis is complicated by the fact that constructors field multiple
cars, and also that constructors come and go, with two 2020 teams
dropping out between years and two joining.  Package idiom to test
this statistically would be to define a combined likelihood function
for both years, keeping track of the year:

<<definecombinedlikelihoodfunction,cache=TRUE>>=
H <- psubs(constructor_2020,"Merc","Merc2020") + psubs(constructor_2021,"Merc","Merc2021")
@

Above, we use generic function \code{psubs()} to change the name of
Mercedes from \code{Merc} to \code{Merc2020} and \code{Merc2021}
respectively.  Note the use of ``\code{+}" to represent addition of
log-likelihoods, corresponding to the assumption of conditional
independence of results.  The null would be simply that the strengths
of \code{Merc2020} and of \code{Merc2021} are identical.  Package
idiom would be to use generic function \code{samep.test()}:

<<usesamep,cache=TRUE>>=
samep.test(H,c("Merc2020","Merc2021"))
@

Above, we see strong evidence for a real decrease in the strength of
the Mercedes team from 2020 to 2021, with $p=0.002$.
